Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-22 paper.tex
progress

Users of type theories often view universe levels as an unwieldy bureaucratic
detail, a necessary annoyance in service of boosting expressive power while
retaining logical consistency. However, universe hierarchies are not going away
any time soon in practical implementations of type theory. In recent
developments of systems, we are getting more universes and more adjacent
features:

\begin{itemize}
\item Agda recently added a limited version cumulativity as an optional feature
  for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
  universe hierarchy to $\omega*2$.
\item Coq added support for cumulative inductive types \cite{timany18cumulative}, and added a form of
  bounded universe polymorphism \cite{ziliani15unification}.
\end{itemize}

\noindent At this point, there is a veritable zoo of universe features in existing
implementations. We have perhaps even more design choices when considering the
formal metatheory of type theories. Do type formers stay in the same
universe, or take the $\sqcup$ of universes of constituent types? Can
eliminators target any universe, or do we instead use lifting operators to cross
levels? What kind of universe polymorphism do we have, can we quantify over
bounds? Is there a type of levels, or are levels in a separate syntactic layer?

@@

However, it would be awkward to pick a particular structure for levels, and
specify a type theory which internalizes that structure; for example
internalizing levels as natural numbers. We do not want to repeat the
specification and semantics for each choice of level structure; instead, we aim
to have a more generic solution.
\begin{enumerate}
  \item We first give a specification of \emph{type theory with dependent
        levels}, or TTDL, where levels and level morphism may depend on typing contexts.
        Here, liftings, universes and type formers are specified, but the internal
        structure of levels is not yet pinned down.
  \item We show that we can extend TTDL with \emph{level reflection} rules, which identify
        levels with particular internal types, thereby getting \emph{type theories with first-class
        levels}, or TTFL.
\end{enumerate}
This decreases the amount of work that we have to do, in order to get semantics
for different level setups. We only need to pick an external level structure
such that it can be also represented using TTDL type formers.

@@

\item
While it is possible to quantify over all levels (using plain $\Pi$ types), it
is not possible to have level polymorphism over all levels. We may try to type
an identity function for all levels, as $\Pi(i : \Lift\,?\,\Lvl^I)(A :
\U\,(\mkLvl\,i)\,?\,?) \to \Lift\,?\,A \to \Lift\,?\,A$. The issue is in
$\U\,(\mkLvl\,i)\,?\,?$, where we have to find a level which is larger than
\emph{every} level. The solution to this issue is to simply add more levels. For
example, for polymorphism over finite levels, we may pick $\omega + \omega$ as
the first limit ordinal which can internally type finite level polymorphism;
this is what Agda 2.6.2 does.
\end{itemize}

@@

Universes:
Martin-Löf (U in U), Girard, Martin-Löf (countable predicative),
CoC, Coq, Luo ECC. Universe poly in Agda (?), in Coq, cumulativity,
Sterling, Coquand normalization (?)

Transfinite: super, higher-order, Mahlo, IR, indexed IR

Harper univ checking?

@@

Second, we only focus on using universes as sized-based classifiers for
types. Stratification features are also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}. It would be interesting to port
universe features in this paper to two-level type theory, as they would
hopefully model a form of stage polymorphism in multi-stage compilation.

>>>>

Users of type theories often view universe levels as an unwieldy bureaucratic
detail, a necessary annoyance in service of boosting expressive power while
retaining logical consistency. However, universe hierarchies are not going away
any time soon in practical implementations of type theory. In recent
developments of systems, we are getting more universes and more adjacent
features:
\begin{itemize}
\item Agda recently added a limited cumulativity as an optional feature
  for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
  universe hierarchy to $\omega*2$.
\item Coq added support for cumulative inductive types \cite{timany18cumulative}
  and a form of bounded universe polymorphism \cite{ziliani15unification}.
\end{itemize}
\noindent At this point, there is a veritable zoo of universe features in existing
implementations. We have perhaps even more design choices when considering the
formal metatheory of type theories. Do type formers stay in the same
universe, or take the $\sqcup$ of universes of constituent types? Can
eliminators target any universe, or do we instead use lifting operators to cross
levels? What kind of universe polymorphism do we have, can we quantify over
bounds? Is there a type of levels, or are levels in a separate syntactic layer?

@@

However, it would be awkward to pick a particular structure for levels, and
specify a type theory which internalizes that structure; for example
internalizing levels as natural numbers. We do not want to repeat the
specification and semantics for each choice of level structure; instead, we aim
to have a more generic solution.
\begin{enumerate}
  \item We first give a specification of \emph{type theory with dependent
        levels}, or TTDL, where levels and level morphisms may depend on typing contexts.
        Here, liftings, universes and type formers are specified, but the internal
        structure of levels is not yet pinned down.
  \item We show that we can extend TTDL with \emph{level reflection} rules, which identify
        levels with particular internal types, thereby getting \emph{type theories with first-class
        levels}, or TTFL.
\end{enumerate}
This decreases the amount of work that we have to do, in order to get semantics
for different level setups. We only need to pick an external level structure
such that it can be also represented using TTDL type formers.

@@

\item
While it is possible to quantify over all levels (using plain $\Pi$ types), it
is not possible to have level polymorphism over all levels. We may try to type
an identity function for all levels, as $\Pi(i : \Lift\,?\,\Lvl^I)(A :
\U\,(\mkLvl\,i)\,?\,?) \to \Lift\,?\,A \to \Lift\,?\,A$. The issue is in
$\U\,(\mkLvl\,i)\,?\,?$, where we would have to find a level which is larger
than \emph{every} level. The solution to this issue is to simply add more
levels. For example, for polymorphism over finite levels, we may pick $\omega +
\omega$ as the first limit ordinal which can internalize finite level
polymorphism; this is what Agda 2.6.2 does.
\end{itemize}

@@

Predicative hierarchies originate as Russell's ramified type theories
\cite{principia}. In the more modern formulations of type theory, Martin-Löf
proposed a countable predicative hierarchy \cite{martinlof73predicative}, as a
way to remedy the inconsistency of the previous version of the theory (which
assumed type-in-type). Harper and Pollack described universe inference with
level assignments and also a form of level polymorphism \cite{harperpollack}.
Sterling \cite{sterling2019algebraic} gave an algebraic specification much like
ours for a type theory with countable cumulative universes, and prove canonicity
for it.

There have been proposals for strengthening universes with various closure
principles and universe operators. Palmgren's super universes and higher-order
universes \cite{Palmgren98onuniverses} and Setzer's Mahlo universes
\cite{setzer00mahlo} are examples for this. These are sufficient to model
transfinite hierarchies, but as we noted in Section
\ref{sec:inductive_recursive_codes}, we do not know how to model strict
inclusions with them. Variants of induction-recursion \cite{dybjer99finite,
  indexedir, positiveir} are particularly flexible and powerful extensions to
universes.

It is worth to summarize here the universe features in the current
type theory implementations.

\textbf{Agda 2.6.1} has $\omega+1$-many non-cumulative predicative universes as
$\Set{i}$, with optional cumulative subtyping only for universes (and not other
type formers) \cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
levels (hence, exluding $\omega$), which supports constructors and some built-in
operations, but no general elimination rule. There is also an $\omega$-bounded
parallel hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda
2.6.2 will extend the hierarchy to $\omega * 2$.

\textbf{Coq 8.13} has $\omega$-many cumulative predicative universes with
cumulative subtyping for all type formers \cite{timany18cumulative}. It supports
bounded universe polymorphism, but it has no internal type for levels, and
universe polymorphic definition are not internally typeable. It also optionally
has an impredicative $\msf{Prop}$ universe and optionally impredicative bottom
$\Seti$ universe. Version 8.13 added experimental support for a parallel
$\omega$-bounded cumulative hierarchy for strict propositions.

\textbf{Lean 3.3} has countable non-cumulative predicative $\msf{Type}_i$
universes with universe polymorphism, and no internal type of levels
\cite{leanmanual}. It also has strict impredicative $\msf{Prop}$.

\textbf{Idris 1} has countable cumulative universes with cumulative subtyping
only for universes, typical-ambiguity-style level inference and no universe
polymorphism \cite{idrisdocs}.

Of the above features, what TTFL does not support is a) impredicativity b) the
interaction of $\msf{Prop}$ and $\msf{Type}$ universes, i.e. the restrictions on
$\msf{Prop}$ elimination.

@@

Second, we only focus on using universes as sized-based classifiers for
types. Stratification features are also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}. It would be interesting to port
universe features in this paper to two-level type theory, as they would
hopefully model a form of stage polymorphism in multi-stage compilation. It
would be nice to try representing $\msf{Prop}$ universes in TTFL as well;
this is closed related to h-level based stratification.
